From Parametricity to Conservation Laws, via Noether's Theorem

### 2013-11-09 mechanics/paper.tex
Final POPL2014 version

\documentclass[preprint]{sigplanconf}

@@

\authorinfo{}{}{}
% \authorinfo{Robert Atkey}
%            {}
%            {bob.atkey@gmail.com}

@@

@@

\begin{abstract}
  Invariance is of paramount importance in programming languages and
  in physics. In programming languages, John Reynolds' theory of
  relational parametricity demonstrates that parametric polymorphic
  programs are invariant under change of data representation, a
  property that yields ``free'' theorems about programs just from
  their types. In physics, Emmy Noether showed that if the action of a
  physical system is invariant under change of coordinates, then the
  physical system has a conserved quantity: a quantity that remains
  constant for all time. Knowledge of conserved quantities can reveal
  deep properties of physical systems. For example, the conservation
  of energy, which by Noether's theorem is a consequence of a system's
  invariance under time-shifting.

@@

\category{CR-number}{subcategory}{third-level}

\terms
term1, term2

\keywords
keyword1, keyword2

@@

Reynolds' theory of relational parametricity \cite{reynolds83types}
tells us that parametrically polymorphic programs automatically
satisfy invariance properties. Such invariance properties are often
called \emph{Free Theorems}, after Wadler \cite{wadler89theorems},
since they follow ``for free'' from the types of programs, rather than
through detailed study of the program text itself. An illustrative
example is the free theorem for programs $f$ with the following type:
\begin{displaymath}
  f : \forall \alpha.~\mathsf{List}~\alpha \to \mathsf{Nat}
\end{displaymath}
Such programs takes lists of $\alpha$s, for any type $\alpha$, to
natural numbers. Using Reynolds' theory of relational parametricity,
Wadler showed that \emph{any} $f$ with this type satisfies the
following property:
\begin{displaymath}
  \forall \alpha, \beta, g : \alpha \to \beta, l : \mathsf{List}~\alpha.~f~[\alpha]~l = f~[\beta]~(\mathrm{map}~g~l)
\end{displaymath}
Thus, any $f$ with the type given above is invariant under mapping
some arbitrary function $g$ over its input. Thinking in terms of
abstract data types and change of data representation, this free
theorem states that $f$ is invariant under change of data
representation from an arbitrary type $\alpha$ to another arbitrary
type $\beta$, via $g$.

@@

Noether's theorem applies to systems described in terms of
\emph{Lagrangians}. For classical mechanics, a Lagrangian is a
function of time, and the positions and velocities of all the
particles in the system, to the difference between the kinetic energy
and the potential energy of the system. For our system with two
particles at positions $x_1$ and $x_2$ with velocities $\dot{x_1}$ and
$\dot{x_2}$ the Lagrangian is:
\begin{equation}\label{eq:coupled-spring-lagrangian}
  L(t,x_1,x_2,\dot{x_1},\dot{x_2}) = \frac{1}{2}m(\dot{x_1}^2 + \dot{x_2}^2) - \frac{1}{2}k(x_1 - x_2)^2
\end{equation}
The first summand represents the total kinetic energy of the system in
terms of the mass and velocities, while the second summand represents
the potential energy contained within the spring due to the distance
between the particles.

From the Lagrangian (\ref{eq:coupled-spring-lagrangian}), using the
\emph{principle of stationary action}, we can derive the following two
equations of motion for this system. (We describe the principle of
stationary action, and the process for deriving the equations of
motion fully in Section~\ref{sec:conservation-laws-from-symmetry}.)
For this system, the equations of motion are a pair of ordinary
differential equations (ODEs) that describe how the positions and
velocities of the particles evolve over time (the second deivative
with respect to time, $\ddot{x_i}$, denotes the acceleration of the
$i$th particle):
\begin{mathpar}
  m\ddot{x_1} = - k(x_1 - x_2)

@@

The invariance property stated in
Equation~(\ref{eq:coupled-spring-translation-invariance}) is highly
reminiscent of the free theorem we stated for the polymorphic program
at the start of this introduction. A change in representation in the
input, whether a change in data representation from a type $\alpha$ to
a type $\beta$, or a change in where time zero is counted from,
results in no change in the output. Since Reynolds' relational
parametricity was allowed us to derive the invariance properties for
the program $f$ just by looking at its type, we ask the following
question:

@@

In this paper, we answer this question positively. We show that it is
indeed the case that we can use a generalised version of Reynolds'
theory of relational parametricity to prove the geometric invariance
properties required for the hypotheses of Noether's theorem. We
construct an extension of System F$\omega$ suitable for writing
invariant Lagrangians, with a relationally parametric model that
allows us to prove invariance properties as free theorems. In our
system, the Lagrangian (\ref{eq:coupled-spring-lagrangian}) we gave
above for describing a system of two particles coupled by a spring
will have the type:
\begin{displaymath}
  \begin{array}{@{}l}
  \forall y \mathord: \mathsf{T}(1).\\
  \quad C^\infty(\mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, y \rangle \times \mathbb{R}\langle 1, y \rangle \times \mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, 0 \rangle, \mathbb{R}\langle 1, 0 \rangle)
\end{array}
\end{displaymath}
We explain this type by breaking it down from left to right. The
quantifier $\forall x \mathord: \mathsf{T}(1)$ indicates that we are
quantifying over all translations $y$ in one-dimensional space, just
as we did implicity in Equation
(\ref{eq:coupled-spring-translation-invariance}). The notation
$C^\infty(-,-)$ denotes the type of \emph{smooth} functions between
spaces. (A smooth function is a function that is infinitely
differentiable.) All the functions that we wish to use as Lagrangians
must be smooth (at least up to second order) for the theory to
work. The type $\mathbb{R}\langle g, x \rangle$ denotes real numbers
that vary with some linear transformation $g$ and translation $x$. In
this case, all the linear transformations are $1$, indicating no
transformation. Likewise, when the translation component is $0$, there
is no translation. Thus a value of type $\mathbb{R}\langle 1, 0
\rangle$ is invariant, but a value of type $\mathbb{R}\langle 1, y
\rangle$ varies with the translation $y$. In the smooth function type
above, in domain space consists of vectors of five real numbers: the
time, the two particle positions, which vary with the translation $y$,
and the two particle velocities.

Atkey, Johann and Kennedy \cite{atkey13abstraction}, building on older
work by Kennedy \cite{kennedy97relational}, have already presented a
polymorphic type system for expressing geometric invariance properties
similar to the translation invariance property in Equation
(\ref{eq:coupled-spring-translation-invariance}). In the present work,
we extend their type system by embedding it within System
F$\omega$. The benefits of this embedding are twofold. Firstly, by
using the richer type-level structure of System F$\omega$, we can
easily add useful indexed types like length indexed vectors and smooth
function types, both of which would have required special treatment in
the special purpose type system presented by Atkey \emph{et
  al.}. Secondly, by using System F$\omega$ as our base, we are able
to reuse the reflexive graph-based relationally parametric semantics
presented by Atkey \cite{atkey12relational}, extending it as
appropriate for our geometric setting.

@@

Our core contributions are threefold:
\begin{enumerate}
\item We reformulate the type system for geometric invariance of
  Atkey, Johann and Kennedy \cite{atkey13abstraction} as an extension
  of System F$\omega$. This shows that the special status of the
  group-indexed types in Atkey \emph{et al.}'s work can be
  incorporated into a standard framework for presenting indexed types.
\item We present a relationally parametric semantics of our extension
  of System F$\omega$. In this model, following Atkey's model for
  basic System F$\omega$ \cite{atkey12relational}, each kind is
  interpreted as a reflexive graph. Our key technical contribution is
  to note that the many geometric groups that we wish to incorporate
  are expressible as groupoids, and hence as reflexive graphs. We
  therefore show that Reynolds' theory of relational parametricity,
  suitably generalised to indexed types by considering reflexive
  graphs, already accomodates geometric invariance properties.
\item Finally, we connect the free theorems derivable from our
  relationally parametric semantics with Noether's theorem for
  deriving conservation laws. We present many examples of invariant
  Lagrangians expressible in our type system, along with the
  consequent conservation laws for physical systems. In each case, the
  necessary invariance property is derived from the type we assign to
  the Lagrangian.
\end{enumerate}

@@

\begin{itemize}
\item In the next section,
  Section~\ref{sec:conservation-laws-from-symmetry}, we describe the
  necessary background to understand Noether's theorem and the
  conservation laws it generates. We briefly introduce the Lagrangian
  formulation of classical mechanics, define precisely what it means
  for a physical system to be invariant under transformations, and
  state Noether's theorem.
\item Having shown how invariance generates conservation laws in
  Section~\ref{sec:conservation-laws-from-symmetry}, in
  Section~\ref{sec:symmetry-from-types}, we show how to derive
  invariance properties from indexed types. We present a relationally
  parametric semantics for System F$\omega$, and show how geometric
  invariance properties can be accomodated within this semantics.
\item In Section~\ref{sec:types-for-classical-mech}, we construct a
  type system for classical mechanics, by extending System F$\omega$
  with the necessary constants for smooth function types and
  combinators for constructing smooth functions. Programming directly
  with our combinators in raw System F$\omega$ is awkward, due to the
  point-free style required, so in Section~\ref{sec:surface-syntax} we
  define a surface syntax for defining smooth functions that is
  translated into our applied System F$\omega$.
\item In Section~\ref{sec:examples} we present several examples of
  using our extension of System F$\omega$ for writing invariant
  Lagrangians that describe physical systems.
\item Section~\ref{sec:conclusions} concludes, and offers directions
  for further work.
\end{itemize}

@@

For Lagrangian mechanics, classical mechanical systems are described
in terms of \emph{Lagrangians} $L = T - V$, where $T$ is the total
kinetic energy of the system, and $V$ is the total potential
energy. The Lagrangian is a function of time, the positions of all the
particles in the system, and their velocities. This is usually written
like so, with abuse of notation for $\dot{q}$ which is a variable
here, not a derivative of a function:
\begin{displaymath}
  L(t, q, \dot{q}) = T - V
\end{displaymath}
In this expression, $q$ and $\dot{q}$ are actually vectors containing
the components of the position and velocity vectors for each
particle. A key feature of Lagrangian mechanics is that the coordinate
system used to describe the system need not necessarily be the usual
cartesian coordinates. For example, one of the components of $q$ might
be the angle of swing of a pendulum. This use of generalised
coordinates is one of the advantages of Lagrangian mechanics over the
Newtonian presentation in terms of positions and forces.

@@

The notation in the Euler-Lagrange equation, while standard, is not
exactly clear, especially to a programming language theorist used to a
more careful treatment of variables. The two partial derivatives
$\frac{\partial L}{\partial \dot{q_i}}$ and $\frac{\partial
  L}{\partial q_i}$ take partial derivatives of $L$ with respect to
$\dot{q_i}$ and $q_i$ \emph{as variables}. The derivative with respect
to time, $\frac{d}{dt}$ then treats $\dot{q_i}$ as a function of time,
hence the appearance of the second derivatives $\ddot{q_i}$ in the
equations of motion. Sussman and Wisdom \cite{sussman01structure}
criticise the standard presentation of Lagrangian mechanics for this
kind of confusing notation, and define a clearer notation based on the
Scheme programming language. We elect to stick with the standard
notation as used in physics textbooks just for familiarity.

@@

The action $\mathcal{S}$ is \emph{invariant} between the endpoints $a$
and $b$, under the transformations $\Phi$ and $\Psi$, if it is the
case that for all paths $q$:
\begin{equation}\label{eq:invariance}
    \int_a^b L(t,q(t),\dot{q}(t)) \mathit{dt}
    = \int_{\Phi(a)}^{\Phi(b)} L(s, q^*(s), \dot{q^*}(s)) \mathit{ds}
\end{equation}
where $q^*(s) = (\Psi \circ q \circ \Phi^{-1})(s)$ is the path $q$
transformed by $\Phi$ and $\Psi$.  By change of variables, the right
hand integral in the above equation is equal to the following
integral:
\begin{displaymath}
  \int_a^b L(\Phi(t), q^*(\Phi(t)), \dot{q^*}(\Phi(t)))\cdot \dot{\Phi}(t) \mathit{dt}
\end{displaymath}
For our application to classical mechanics, the endpoints $a$ and $b$
are arbitrary, so we can rewrite Equation (\ref{eq:invariance}) to the
following equation between applications of the Lagrangian:
\begin{equation}\label{eq:simplified-invariance}
  L(t,q(t),\dot{q}(t)) = L(\Phi(t), q^*(\Phi(t)), \dot{q^*}(\Phi(t)))\cdot \dot{\Phi}(t)
\end{equation}
For most examples, it will be the case that $\Phi$ has the form
$\Phi(t) = t + t'$ and $\Psi$ is an affine transformation $\Psi(q)
= Gq + x$, where $G$ is an $n\mathord\times n$ real matrix and $x \in
\mathbb{R}^n$. In this case, Equation (\ref{eq:simplified-invariance})
simplifies further, using the definition of $q^*$ and the chain rule
of differentiation, to:
\begin{equation}
  \label{eq:simple-invariance}
  L(t,q(t),\dot{q}(t)) = L(t + t', Gq(t) + x, G\dot{q}(t))
\end{equation}
Equation (\ref{eq:simple-invariance}) is obviously implied by the
following equation, where we treat $q$ and $\dot{q}$ as universally
quantified variables, instead of as functions of time:
\begin{equation}
  \label{eq:simplest-invariance}
  L(t,q,\dot{q}) = L(t + t', Gq + x, G\dot{q})
\end{equation}
Equation (\ref{eq:simplest-invariance}) is the form of invariance that
the free theorems we derive from the types of our typed language for
classical mechanics will usually take. We define our type system for
classical mechanics in Sections~\ref{sec:symmetry-from-types}
and~\ref{sec:types-for-classical-mech} below.

\begin{example}\label{ex:invariance-examples}
  In the introduction, we stated that the Lagrangian
  (\ref{eq:coupled-spring-lagrangian}) describing the spring-coupled
  system is invariant under translation in space. Our running example
  in this section, the Lagrangian (\ref{eq:simple-g-lagrangian}) for
  the single particle under gravity, is not invariant under all
  spatial translations, due to the reference in the potential energy
  term to the distance from the fixed baseline at $y = 0$. It is
  invariant under translations along the $x$-axis, however. Both
  Lagrangians are also invariant in translations in time, a special
  case of Equation~(\ref{eq:simplest-invariance}):
  \begin{displaymath}
    L(t,q,\dot{q}) = L(t+t',q,\dot{q})
  \end{displaymath}
  This invariance property is simply a consequence of the fact that
  neither Lagrangian explicitly mentions the time $t$.
\end{example}

@@

\begin{theorem}[Noether]
  If the action
  \begin{displaymath}
    \mathcal{S}[q;a;b] = \int_a^b L(t,q(t),\dot{q}(t))~dt
  \end{displaymath}
  is invariant under $\epsilon$-indexed families of transformations
  $\Phi_\epsilon$ and $\Psi_\epsilon$, then the equation
  \begin{equation}
    \label{eq:conservation-law}
    \frac{d}{dt}\left(\sum_{i=1}^n \frac{\partial L}{\partial \dot{q_i}}\psi_i + \left(L - \sum_{i=1}^n \dot{q_i}\frac{\partial L}{\partial \dot{q_i}}\right) \phi \right) = 0
  \end{equation}
  holds for all paths $q$ satisfying the Euler-Lagrange equations
  (\ref{eq:euler-lagrange}), where $\psi = \left. \frac{\partial
      \Psi}{\partial \epsilon} \right|_{\epsilon = 0}$ and $\phi =
  \left. \frac{\partial \Phi}{\partial \epsilon} \right|_{\epsilon =
    0}$.
\end{theorem}

@@

@@

We deviate from the standard presentation of System F$\omega$ in that
we distinguish between large and small kinds, and only allow
quantification over small kinds. In particular, the kind of types $*$
is \emph{not} small. The reason for this is that we desire a simple
set-theoretic model. However, requiring that we have a set of sets
closed under large products, in order to interpret quantification over
the kind $*$, would force us to use an intuitionistic metatheory
\cite{pitts87polymorphism}. The basic results of the calculus of
variations that we are relying on, such as the derivation of the
Euler-Lagrange equations, use excluded middle in their proofs, so an
intuitionistic metatheory is not an option. We therefore just assume
that we have a set of ``small'' sets, which includes the real numbers
and is closed under subsets, products, function space and
comprehension.

\paragraph{Syntax} The kinds of System F$\omega$ include at least the
base kind of types, $*$, and function and product kinds, as generated
by the following grammar:
\begin{displaymath}
  \kappa ::= * \sepbar \kappa_1 \to \kappa_2 \sepbar \kappa_1 \times \kappa_2 \sepbar \cdots
\end{displaymath}
We will extend the collection of kinds in
Section~\ref{sec:discrete-kinds} with type-level natural numbers, and
in Section~\ref{sec:groupoid-kinds}

@@

The well-kinded types, type equalities, and well-typed terms of System
F$\omega$ are shown in Figures~\ref{fig:types},
\ref{fig:type-equality} and~\ref{fig:terms}, respectively. Note that
typing contexts $\Gamma$ only contain well-kinded types of kind
$*$. Terms also have an equational theory, with $\beta\eta$-laws for
functions, products, and universal quantification, which we omit. Our
presentation of System F$\omega$ is entirely standard (see, e.g.,
Pierce~\cite{pierce02types}), except for the restriction to small
kinds in the formation of universal kinds $\forall
\alpha\mathord:\kappa. A$. As is the case for kinds, we will extend
the types, type equalities and terms in
Sections~\ref{sec:discrete-kinds},~\ref{sec:groupoid-kinds}
and~\ref{sec:types-for-classical-mech}, below.

>>>>

\documentclass{sigplanconf}

\pdfpagewidth=8.5in
\pdfpageheight=11in

@@

\authorinfo{Robert Atkey}{}{bob.atkey@gmail.com}

@@

\exclusivelicense
\conferenceinfo{POPL~'14}{January 22--24, 2014, San Diego, CA, USA}
\copyrightyear{2014}
\copyrightdata{978-1-4503-2544-8/14/01}
\doi{2535838.2535867}

@@

\begin{abstract}
  Invariance is of paramount importance in programming languages and
  in physics. In programming languages, John Reynolds' theory of
  relational parametricity demonstrates that parametric polymorphic
  programs are invariant under change of data representation, a
  property that yields ``free'' theorems about programs just from
  their types. In physics, Emmy Noether showed that if the action of a
  physical system is invariant under change of coordinates, then the
  physical system has a conserved quantity: a quantity that remains
  constant for all time. Knowledge of conserved quantities can reveal
  deep properties of physical systems. For example, the conservation
  of energy is by Noether's theorem a consequence of a system's
  invariance under time-shifting.

@@

\category{D.3.1}{Programming Languages}{Formal Definitions and Theory}
\category{J.2}{Physical Sciences and Engineering}{Physics}

\terms
  Languages, Theory, Types, Physics

\keywords
  classical mechanics; relational parametricity; higher-kinded types; invariance; conservation laws

@@

Reynolds' theory of relational parametricity \cite{reynolds83types}
tells us that parametrically polymorphic programs automatically
satisfy invariance properties. Such invariance properties are often
called \emph{Free Theorems}, after Wadler \cite{wadler89theorems},
since they follow ``for free'' from the types of programs, rather than
through detailed study of the program text. An illustrative example is
the free theorem for programs $f$ with the following type:
\begin{displaymath}
  f : \forall \alpha.~\mathsf{List}~\alpha \to \mathsf{Nat}
\end{displaymath}
Such programs take lists of $\alpha$s, for any type $\alpha$, and
return natural numbers. Using Reynolds' theory of relational
parametricity, Wadler showed that \emph{any} $f$ with this type
satisfies the following property:
\begin{displaymath}
  \forall \alpha, \beta, g : \alpha \to \beta, l : \mathsf{List}~\alpha.~f~[\alpha]~l = f~[\beta]~(\mathrm{map}~g~l)
\end{displaymath}
Thus, any $f$ with the type given above is invariant under mapping
some arbitrary function $g$ over its input. Thinking in terms of
abstract data types and change of data representation, this free
theorem states that $f$ is invariant under change of data
representation from an arbitrary type $\alpha$ to another arbitrary
type $\beta$, via $g$.

@@

Noether's theorem applies to systems described in terms of
\emph{Lagrangians}. For classical mechanics, a Lagrangian is a
function of time and the positions and velocities of all the
particles in the system, to the difference between the kinetic energy
and the potential energy of the system. For our system with two
particles at positions $x_1$ and $x_2$ with velocities $\dot{x_1}$ and
$\dot{x_2}$ the Lagrangian is:
\begin{equation}\label{eq:coupled-spring-lagrangian}
  L(t,x_1,x_2,\dot{x_1},\dot{x_2}) = \frac{1}{2}m(\dot{x_1}^2 + \dot{x_2}^2) - \frac{1}{2}k(x_1 - x_2)^2
\end{equation}
The first summand represents the total kinetic energy of the system in
terms of the mass and velocities, while the second summand represents
the potential energy contained within the spring due to the distance
between the particles.

From the Lagrangian (\ref{eq:coupled-spring-lagrangian}), using the
\emph{principle of stationary action}, we can derive the following two
equations of motion for this system. (We describe the principle of
stationary action, and the process for deriving the equations of
motion fully in Section~\ref{sec:conservation-laws-from-symmetry}.)
For this system, the equations of motion are a pair of ordinary
differential equations (ODEs) that describe how the positions and
velocities of the particles evolve over time (the second derivative
with respect to time, $\ddot{x_i}$, denotes the acceleration of the
$i$th particle):
\begin{mathpar}
  m\ddot{x_1} = - k(x_1 - x_2)

@@

The invariance property stated in
Equation~(\ref{eq:coupled-spring-translation-invariance}) is highly
reminiscent of the free theorem we stated for the polymorphic program
at the start of this introduction. A change in representation in the
input, whether a change in data representation from a type $\alpha$ to
a type $\beta$, or a change in where time zero is counted from,
results in no change in the output. Since Reynolds' relational
parametricity allowed us to derive the invariance properties for the
program $f$ just by looking at its type, we ask the following
question:

@@

In this paper, we answer this question positively. We show that it is
indeed the case that we can use a generalised version of Reynolds'
theory of relational parametricity to prove the geometric invariance
properties required for the hypotheses of Noether's theorem. We
construct an extension of System F$\omega$ suitable for writing
invariant Lagrangians, with a relationally parametric model that
allows us to prove invariance properties as free theorems. In our
system, the Lagrangian (\ref{eq:coupled-spring-lagrangian}) we gave
above for describing a system of two particles coupled by a spring
will have the type:
\begin{displaymath}
  \begin{array}{@{}l}
  \forall y \mathord: \mathsf{T}(1).\\
  \quad C^\infty(\mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, y \rangle \times \mathbb{R}\langle 1, y \rangle \times \mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, 0 \rangle, \mathbb{R}\langle 1, 0 \rangle)
\end{array}
\end{displaymath}
We explain this type by breaking it down from left to right. The
quantifier $\forall y \mathord: \mathsf{T}(1)$ indicates that we are
quantifying over all translations $y$ in one-dimensional space, just
as we did implicitly in Equation
(\ref{eq:coupled-spring-translation-invariance}). The notation
$C^\infty(-,-)$ denotes the type of \emph{smooth} functions between
spaces. (A smooth function is a function that is infinitely
differentiable.) All the functions that we wish to use as Lagrangians
must be smooth (at least up to second order) for the theory to
work. The type $\mathbb{R}\langle g, x \rangle$ denotes real numbers
that vary with some linear transformation $g$ and translation $x$. In
this case, all the linear transformations are $1$, indicating no
transformation. Likewise, when the translation component is $0$, there
is no translation. Thus a value of type $\mathbb{R}\langle 1, 0
\rangle$ is invariant, but a value of type $\mathbb{R}\langle 1, y
\rangle$ varies with the translation $y$. In the smooth function type
above, the domain space consists of vectors of five real numbers: the
time, the two particle positions, which vary with the translation $y$,
and the two particle velocities.

Atkey, Johann and Kennedy \cite{atkey13abstraction}, building on
work by Kennedy \cite{kennedy97relational}, have already presented a
polymorphic type system for expressing geometric invariance properties
similar to the translation invariance property in Equation
(\ref{eq:coupled-spring-translation-invariance}). In the present work,
we extend their type system by embedding it within System
F$\omega$. The benefits of this embedding are twofold. Firstly, by
using the richer type-level structure of System F$\omega$, we can
easily add useful indexed types like length indexed vectors and smooth
function types, both of which would have required special treatment in
the special purpose type system presented by Atkey \emph{et
  al.}. Secondly, by using System F$\omega$ as our base, we are able
to reuse the reflexive graph-based relationally parametric semantics
presented by Atkey \cite{atkey12relational}, extending it as
appropriate for our geometric setting.

@@

Our core contributions are threefold:
\begin{enumerate}
\item We reformulate the type system for geometric invariance of
  Atkey, Johann and Kennedy \cite{atkey13abstraction} as an extension
  of System F$\omega$. This shows that the special status of the
  group-indexed types in Atkey \emph{et al.}'s work can be
  incorporated into a standard framework for indexed types.
\item We present a relationally parametric semantics of our extension
  of System F$\omega$. In this model, following Atkey's model for
  basic System F$\omega$ \cite{atkey12relational}, each kind is
  interpreted as a reflexive graph. Our key technical contribution is
  to note that the many geometric groups that we wish to incorporate
  are expressible as groupoids, and hence as reflexive graphs. We
  therefore show that Reynolds' theory of relational parametricity,
  suitably generalised to indexed types by considering reflexive
  graphs, already accommodates geometric invariance properties.
\item Finally, we connect the free theorems derivable from our
  relationally parametric semantics with Noether's theorem for
  deriving conservation laws. We present many examples of invariant
  Lagrangians expressible in our type system, along with the
  consequent conservation laws for physical systems. In each case, the
  necessary invariance property is derived from the type of the
  Lagrangian.
\end{enumerate}

@@

\begin{itemize}
\item In the next section,
  Section~\ref{sec:conservation-laws-from-symmetry}, we describe the
  necessary background to understand Noether's theorem and the
  conservation laws it generates. We briefly introduce the Lagrangian
  formulation of classical mechanics, define precisely what it means
  for a physical system to be invariant under transformations, and
  state Noether's theorem.
\item Having shown how invariance generates conservation laws in
  Section~\ref{sec:conservation-laws-from-symmetry}, in
  Section~\ref{sec:symmetry-from-types} we show how to derive
  invariance properties from indexed types. We present a relationally
  parametric semantics for System F$\omega$, and show how geometric
  invariance properties can be accomodated within this semantics.
\item In Section~\ref{sec:types-for-classical-mech} we construct a
  type system for classical mechanics by extending System F$\omega$
  with the necessary constants for smooth function types and
  combinators for constructing smooth functions. Programming directly
  with our combinators in raw System F$\omega$ is awkward, due to the
  point-free style required, so in Section~\ref{sec:surface-syntax} we
  define a surface syntax for defining smooth functions that is
  translated into our applied System F$\omega$.
\item In Section~\ref{sec:examples} we present several examples of
  using our extension of System F$\omega$ for writing invariant
  Lagrangians that describe physical systems.
\item Section~\ref{sec:conclusions} concludes, and offers directions
  for further work.
\end{itemize}

@@

For Lagrangian mechanics, classical mechanical systems are described
in terms of \emph{Lagrangians} $L = T - V$, where $T$ is the total
kinetic energy of the system, and $V$ is the total potential
energy. The Lagrangian is a function of time, the positions of all the
particles in the system, and their velocities. This is usually written
like so, with abuse of notation for $\dot{q}$ which is a variable
here, not the derivative of a function:
\begin{displaymath}
  L(t, q, \dot{q}) = T - V
\end{displaymath}
In this expression, $q$ and $\dot{q}$ are actually vectors containing
the components of the position and velocity vectors for each
particle. A key feature of Lagrangian mechanics is that the coordinate
system used to describe the system need not necessarily be the usual
cartesian coordinates. For example, one of the components of $q$ might
be the angle of swing of a pendulum. This use of generalised
coordinates is one of the advantages of Lagrangian mechanics over the
Newtonian presentation in terms of positions and forces.

@@

The notation in the Euler-Lagrange equation, while standard, is not
exactly clear, especially to a programming language theorist used to a
more careful treatment of variables. The two partial derivatives
$\frac{\partial L}{\partial \dot{q_i}}$ and $\frac{\partial
  L}{\partial q_i}$ take partial derivatives of $L$ with respect to
$\dot{q_i}$ and $q_i$ \emph{as variables}. The derivative with respect
to time, $\frac{d}{dt}$, then treats $\dot{q_i}$ as a function of
time, hence the appearance of the second derivatives $\ddot{q_i}$ in
the equations of motion. Sussman and Wisdom \cite{sussman01structure}
criticise the standard presentation of Lagrangian mechanics for this
kind of confusing notation, and define a clearer notation based on the
Scheme programming language. We elect to stick with the standard
notation as used in physics textbooks just for familiarity.

@@

The action $\mathcal{S}$ is \emph{invariant} between the endpoints $a$
and $b$, under the transformations $\Phi$ and $\Psi$, if it is the
case that for all paths $q$:
\begin{equation}\label{eq:invariance}
    \int_a^b L(t,q(t),\dot{q}(t)) \mathit{dt}
    = \int_{\Phi(a)}^{\Phi(b)} L(s, q^*(s), \frac{d q^*}{ds}(s)) \mathit{ds}
\end{equation}
where $q^*(s) = (\Psi \circ q \circ \Phi^{-1})(s)$ is the path $q$
transformed by $\Phi$ and $\Psi$.  By change of variables, the right
hand integral in the above equation is equal to the following
integral:
\begin{displaymath}
  \int_a^b L(\Phi(t), q^*(\Phi(t)), \frac{d q^*}{ds}(\Phi(t)))\cdot \abs{\dot{\Phi}(t)} \mathit{dt}
\end{displaymath}
For our application to classical mechanics, the endpoints $a$ and $b$
are arbitrary, so we can rewrite Equation (\ref{eq:invariance}) to the
following equation between applications of the Lagrangian:
\begin{equation}\label{eq:simplified-invariance}
  L(t,q(t),\dot{q}(t)) = L(\Phi(t), q^*(\Phi(t)), \frac{d q^*}{ds}(\Phi(t)))\cdot \abs{\dot{\Phi}(t)}
\end{equation}
For most examples, it will be the case that $\Phi$ has the form
$\Phi(t) = t + t'$ and $\Psi$ is an affine transformation $\Psi(q)
= Gq + x$, where $G$ is an $n\mathord\times n$ real matrix and $x \in
\mathbb{R}^n$. In this case, Equation (\ref{eq:simplified-invariance})
simplifies further, using the definition of $q^*$ and the chain rule
of differentiation, to:
\begin{equation}
  \label{eq:simple-invariance}
  L(t,q(t),\dot{q}(t)) = L(t + t', Gq(t) + x, G\dot{q}(t))
\end{equation}
Equation (\ref{eq:simple-invariance}) is obviously implied by the
following equation, where we treat $q$ and $\dot{q}$ as universally
quantified variables, instead of as functions of time:
\begin{equation}
  \label{eq:simplest-invariance}
  L(t,q,\dot{q}) = L(t + t', Gq + x, G\dot{q})
\end{equation}
Equation (\ref{eq:simplest-invariance}) is the form of invariance that
the free theorems we derive from the types of our typed language for
classical mechanics will usually take. We define our type system for
classical mechanics in Sections~\ref{sec:symmetry-from-types}
and~\ref{sec:types-for-classical-mech} below.

\begin{example}\label{ex:invariance-examples}
  In the introduction, we stated that the Lagrangian
  (\ref{eq:coupled-spring-lagrangian}) describing the spring-coupled
  system is invariant under translation in space. Our running example
  in this section, the Lagrangian (\ref{eq:simple-g-lagrangian}) for
  the single particle under gravity, is not invariant under all
  spatial translations, due to the reference in the potential energy
  term to the distance from the fixed baseline at $y = 0$. It is
  invariant under translations along the $x$-axis, however. Both
  Lagrangians are also invariant under translations in time, a special
  case of Equation~(\ref{eq:simplest-invariance}):
  \begin{displaymath}
    L(t,q,\dot{q}) = L(t+t',q,\dot{q})
  \end{displaymath}
  This invariance property is simply a consequence of the fact that
  neither Lagrangian explicitly mentions the time $t$.
\end{example}

@@

\begin{theorem}[Noether]\label{thm:noether}
  If the action
  \begin{displaymath}
    \mathcal{S}[q;a;b] = \int_a^b L(t,q(t),\dot{q}(t))~dt
  \end{displaymath}
  is invariant under $\epsilon$-indexed families of transformations
  $\Phi_\epsilon$ and $\Psi_\epsilon$, then the equation
  \begin{equation}
    \label{eq:conservation-law}
    \frac{d}{dt}\left(\sum_{i=1}^n \frac{\partial L}{\partial \dot{q_i}}\psi_i + \left(L - \sum_{i=1}^n \dot{q_i}\frac{\partial L}{\partial \dot{q_i}}\right) \phi \right) = 0
  \end{equation}
  holds for all paths $q$ satisfying the Euler-Lagrange equations
  (\ref{eq:euler-lagrange}), where $\psi = \left. \frac{\partial
      \Psi}{\partial \epsilon} \right|_{\epsilon = 0}$ and $\phi =
  \left. \frac{\partial \Phi}{\partial \epsilon} \right|_{\epsilon =
    0}$.
\end{theorem}

@@

% Theorem \ref{thm:noether} is a special case of what is known as
% Noether's ``first theorem''. The general case of Theorem
% \ref{thm:noether} applies to situations where the action is an
% integral over regions of $n$-dimensional space, and the
% transformations $\Psi$ can also depend on time. Noether's second
% theorem generalises further to the case when there are infinitely many
% parameters $\epsilon$. This is the case when considering local
% symmetries that vary at different points in spacetime.

@@

We deviate from the standard presentation of System F$\omega$ in that
we distinguish between large and small kinds, and only allow
quantification over small kinds. In particular, the kind of types $*$
is \emph{not} small. The reason for this choice is that we desire a
simple set-theoretic model. However, requiring that we have a set of
sets closed under large products, in order to interpret quantification
over the kind $*$, would force us to use an intuitionistic metatheory
\cite{pitts87polymorphism}. The basic results of the Calculus of
Variations that we are relying on, such as the derivation of the
Euler-Lagrange equations, use excluded middle in their proofs, so an
intuitionistic metatheory is not an option. We therefore just assume
that we have a set of ``small'' sets, which includes the real numbers
and is closed under subsets, products, function spaces and
set comprehensions.

\paragraph{Syntax} The kinds of System F$\omega$ include at least the
base kind of types, $*$, and function and product kinds, as generated
by the following grammar:
\begin{displaymath}
  \kappa ::= * \sepbar \kappa_1 \to \kappa_2 \sepbar \kappa_1 \times \kappa_2 \sepbar \cdots
\end{displaymath}
We will extend the collection of kinds in
Section~\ref{sec:discrete-kinds} with type-level natural numbers, and
in Section~\ref{sec:groupoid-kinds} with groupoid kinds.

@@

The well-kinded types, type equalities, and well-typed terms of System
F$\omega$ are shown in Figures~\ref{fig:types},
\ref{fig:type-equality} and~\ref{fig:terms}, respectively. Kinding
contexts $\alpha_1 : \kappa_1, ..., \alpha_n : \kappa_n$ are denoted
by $\Theta$ and typing contexts $x_1 : A_1, ..., x_n : A_n$ are
denoted by $\Gamma$. Note that typing contexts $\Gamma$ only contain
well-kinded types of kind $*$. Terms also have an equational theory,
with $\beta\eta$-laws for functions, products, and universal
quantification, which we omit. Our presentation of System F$\omega$ is
entirely standard (see, e.g., Pierce~\cite{pierce02types}), except for
the restriction to small kinds in the formation of universal kinds
$\forall \alpha\mathord:\kappa. A$. As is the case for kinds, we will
extend the types, type equalities, and terms in
Sections~\ref{sec:discrete-kinds},~\ref{sec:groupoid-kinds}
and~\ref{sec:types-for-classical-mech}, below.
